diffEnrich: An \proglang{R} Package to Compare Functional Enrichment Between Two Experimentally-derived Groups of Genes by Connecting to the KEGG REST API

### 2019-11-23 jss-article-rnw-diffEnrich/article.Rnw
First draft jss article

@@

\begin{table}[H]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \tiny
  \begin{adjustbox}{width=\textwidth}
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

>>>>

\subsection{\code{plotFoldEnrichment}}

\code{plotFoldEnrichment} generates a grouped bar plot using
\code{ggplot2} (\cite{Hadley:2106}) and the \code{ggnewscale}
package (\cite{Elio:2019}). There are 3 arguments: 1)
\code{de_res} is the dataframe generated from the
\code{diffEnrich} function, 2) \code{pval} is the threshold
for the adjusted p-value associated with differential
enrichment that will filter which KEGG pathways to plot, and
3) after filtering based on \code{pval}, \code{N} tells the
function how many pathways to plot. It is important to make
a note that the significance of the fold change is associated
with the number of genes in the gene list. Notice that in this
example the pathways in gene list 2 have smaller fold changes
(shorter bars) than those in list 1, but that many of them are
more significant (darker blue). This is because there are
more genes in gene list 2 compared to gene list 1.

\begin{figure}[H]
% \centering
<<visualization_2, echo=TRUE, message=FALSE, out.height='4in', out.width='5.0in'>>=
## Plot fold enrichment
plotFoldEnrichment(de_res = diff_enrich, pval = 0.05, N = 5)
@
\caption{\label{fig:barchart} \textbf{Example of a differential enrichment graphic.}
KEGG pathways are plotted on the y-axis and fold enrichment is
plotted on the x-axis. Each KEGG pathway has a bar depicting its
fold enrichment in list 1 (red) and its fold enrichment in list 2
(blue). The transparency of the bars correspond to the unadjusted
p-value for the pathway's enrichment in the given list. The p-value
presented as text to the right of each pair of bars is the adjusted
p-value (user defined: default is FDR) associated with the
differential enrichment of the pathway between the two lists, and
the pathways are ordered from top to bottom by this p-value (i.e.
smallest p-value on top of plot, and largest p-value on bottom of
plot). The dotted line represents no enrichment (i.e a fold enrichment of 1).
The number of genes used for analysis from each gene list (recall
that this number may not be the same as the number of genes in
the user’s original list) are reported below their respective
p-values in the legend. In this example, all five pathways are
differentially enriched with more enrichment in List 1 than in
List 2.}
\end{figure}

@@

\begin{table}[H]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \begin{adjustbox}{width=\textwidth, angle=90}
  \fontsize{90}{125}\selectfont
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

<<diffEnrich_head, echo = FALSE, eval = FALSE>>=
suptb2 <- head(diff_enrich$de_table)
xtable(suptb2)
@

\begin{table}[H]
\centering
  \caption{Supplementary Table 2. First 6 results generated by diffEnrich}
  \begin{adjustbox}{width=\textwidth, angle=90}
  \fontsize{50}{80}\selectfont
\begin{tabular}{rllrrrrrrrrrrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_DATABASE\_cnt & KEGG\_PATHWAY\_in\_list1 & KEGG\_DATABASE\_in\_list1 & expected\_list1 & enrich\_p\_list1 & p\_adj\_list1 & fold\_enrichment\_list1 & KEGG\_PATHWAY\_in\_list2 & KEGG\_DATABASE\_in\_list2 & expected\_list2 & enrich\_p\_list2 & p\_adj\_list2 & fold\_enrichment\_list2 & odd\_ratio & diff\_enrich\_p & diff\_enrich\_adjusted \\
  \hline
rno04530 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 8856 & 19.00 & 295 & 5.66 & 0.00 & 0.00 & 3.36 & 131.00 & 5308 & 101.89 & 0.00 & 0.00 & 1.29 & 0.37 & 0.00 & 0.00 \\
  rno05135 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 8856 & 16.00 & 295 & 4.26 & 0.00 & 0.00 & 3.75 & 105.00 & 5308 & 76.72 & 0.00 & 0.00 & 1.37 & 0.35 & 0.00 & 0.00 \\
  rno05210 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 8856 & 12.00 & 295 & 2.93 & 0.00 & 0.00 & 4.09 & 81.00 & 5308 & 52.74 & 0.00 & 0.00 & 1.54 & 0.37 & 0.00 & 0.00 \\
  rno05213 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 8856 & 9.00 & 295 & 1.93 & 0.00 & 0.00 & 4.66 & 55.00 & 5308 & 34.76 & 0.00 & 0.00 & 1.58 & 0.33 & 0.01 & 0.01 \\
  rno04660 & rno04660 & T cell receptor signaling pathway - Rattus norvegicus (rat) & 106 & 8856 & 11.00 & 295 & 3.53 & 0.00 & 0.01 & 3.12 & 79.00 & 5308 & 63.53 & 0.00 & 0.00 & 1.24 & 0.39 & 0.01 & 0.01 \\
  rno04657 & rno04657 & IL-17 signaling pathway - Rattus norvegicus (rat) &  95 & 8856 & 9.00 & 295 & 3.16 & 0.00 & 0.03 & 2.84 & 59.00 & 5308 & 56.94 & 0.37 & 0.46 & 1.04 & 0.36 & 0.01 & 0.01 \\
   \hline
\end{tabular}
\end{adjustbox}
\end{table}

### 2019-11-23 jss-article-rnw-diffEnrich/article.tex
First draft jss article

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Kegg Release: Release\_92.0+\_11-22\_Nov\_19}}\end{kframe}
\end{knitrout}

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# ncbi\_to\_kegg2019-11-22Release\_92.0+\_11-22\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# kegg\_to\_pathway2019-11-22Release\_92.0+\_11-22\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# pathway\_to\_species2019-11-22Release\_92.0+\_11-22\_Nov\_19.txt}}

@@

@@

\begin{table}[H]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \tiny
  \begin{adjustbox}{width=\textwidth}
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

>>>>

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Kegg Release: Release\_92.0+\_11-23\_Nov\_19}}\end{kframe}
\end{knitrout}

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# ncbi\_to\_kegg2019-11-22Release\_92.0+\_11-23\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# kegg\_to\_pathway2019-11-22Release\_92.0+\_11-23\_Nov\_19.txt}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# pathway\_to\_species2019-11-22Release\_92.0+\_11-23\_Nov\_19.txt}}

@@

\subsection{\code{plotFoldEnrichment}}

\code{plotFoldEnrichment} generates a grouped bar plot using
\code{ggplot2} (\cite{Hadley:2106}) and the \code{ggnewscale}
package (\cite{Elio:2019}). There are 3 arguments: 1)
\code{de_res} is the dataframe generated from the
\code{diffEnrich} function, 2) \code{pval} is the threshold
for the adjusted p-value associated with differential
enrichment that will filter which KEGG pathways to plot, and
3) after filtering based on \code{pval}, \code{N} tells the
function how many pathways to plot. It is important to make
a note that the significance of the fold change is associated
with the number of genes in the gene list. Notice that in this
example the pathways in gene list 2 have smaller fold changes
(shorter bars) than those in list 1, but that many of them are
more significant (darker blue). This is because there are
more genes in gene list 2 compared to gene list 1.

\begin{figure}[H]
% \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Plot fold enrichment}
\hlkwd{plotFoldEnrichment}\hlstd{(}\hlkwc{de_res} \hlstd{= diff_enrich,} \hlkwc{pval} \hlstd{=} \hlnum{0.05}\hlstd{,} \hlkwc{N} \hlstd{=} \hlnum{5}\hlstd{)}
\end{alltt}
\end{kframe}
\includegraphics[width=5.0in,height=4in]{figure/visualization_2-1}

\end{knitrout}
\caption{\label{fig:barchart} \textbf{Example of a differential enrichment graphic.}
KEGG pathways are plotted on the y-axis and fold enrichment is
plotted on the x-axis. Each KEGG pathway has a bar depicting its
fold enrichment in list 1 (red) and its fold enrichment in list 2
(blue). The transparency of the bars correspond to the unadjusted
p-value for the pathway's enrichment in the given list. The p-value
presented as text to the right of each pair of bars is the adjusted
p-value (user defined: default is FDR) associated with the
differential enrichment of the pathway between the two lists, and
the pathways are ordered from top to bottom by this p-value (i.e.
smallest p-value on top of plot, and largest p-value on bottom of
plot). The dotted line represents no enrichment (i.e a fold enrichment of 1).
The number of genes used for analysis from each gene list (recall
that this number may not be the same as the number of genes in
the user’s original list) are reported below their respective
p-values in the legend. In this example, all five pathways are
differentially enriched with more enrichment in List 1 than in
List 2.}
\end{figure}

@@

\begin{table}[H]
  \centering
  \caption{Supplementary Table 1. First 6 results generated by pathEnrich.}
  \begin{adjustbox}{width=\textwidth, angle=90}
  \fontsize{90}{125}\selectfont
  \begin{tabular}{rllrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_PATHWAY\_in\_list & KEGG\_DATABASE\_cnt & KEG\_DATABASE\_in\_list & expected & enrich\_p & p\_adj & fold\_enrichment \\
  \hline
95 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 19.00 & 8856 & 295 & 5.66 & 0.00 & 0.00 & 3.36 \\
  172 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 16.00 & 8856 & 295 & 4.26 & 0.00 & 0.00 & 3.75 \\
  194 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 12.00 & 8856 & 295 & 2.93 & 0.00 & 0.00 & 4.09 \\
  212 & rno05231 & Choline metabolism in cancer - Rattus norvegicus (rat) &  99 & 12.00 & 8856 & 295 & 3.30 & 0.00 & 0.00 & 3.64 \\
  197 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 9.00 & 8856 & 295 & 1.93 & 0.00 & 0.00 & 4.66 \\
  66 & rno04144 & Endocytosis - Rattus norvegicus (rat) & 275 & 22.00 & 8856 & 295 & 9.16 & 0.00 & 0.00 & 2.40 \\
   \hline
  \end{tabular}
\end{adjustbox}
\end{table}

\begin{table}[H]
\centering
  \caption{Supplementary Table 2. First 6 results generated by diffEnrich}
  \begin{adjustbox}{width=\textwidth, angle=90}
  \fontsize{50}{80}\selectfont
\begin{tabular}{rllrrrrrrrrrrrrrrrrr}
  \hline
 & KEGG\_PATHWAY\_ID & KEGG\_PATHWAY\_description & KEGG\_PATHWAY\_cnt & KEGG\_DATABASE\_cnt & KEGG\_PATHWAY\_in\_list1 & KEGG\_DATABASE\_in\_list1 & expected\_list1 & enrich\_p\_list1 & p\_adj\_list1 & fold\_enrichment\_list1 & KEGG\_PATHWAY\_in\_list2 & KEGG\_DATABASE\_in\_list2 & expected\_list2 & enrich\_p\_list2 & p\_adj\_list2 & fold\_enrichment\_list2 & odd\_ratio & diff\_enrich\_p & diff\_enrich\_adjusted \\
  \hline
rno04530 & rno04530 & Tight junction - Rattus norvegicus (rat) & 170 & 8856 & 19.00 & 295 & 5.66 & 0.00 & 0.00 & 3.36 & 131.00 & 5308 & 101.89 & 0.00 & 0.00 & 1.29 & 0.37 & 0.00 & 0.00 \\
  rno05135 & rno05135 & Yersinia infection - Rattus norvegicus (rat) & 128 & 8856 & 16.00 & 295 & 4.26 & 0.00 & 0.00 & 3.75 & 105.00 & 5308 & 76.72 & 0.00 & 0.00 & 1.37 & 0.35 & 0.00 & 0.00 \\
  rno05210 & rno05210 & Colorectal cancer - Rattus norvegicus (rat) &  88 & 8856 & 12.00 & 295 & 2.93 & 0.00 & 0.00 & 4.09 & 81.00 & 5308 & 52.74 & 0.00 & 0.00 & 1.54 & 0.37 & 0.00 & 0.00 \\
  rno05213 & rno05213 & Endometrial cancer - Rattus norvegicus (rat) &  58 & 8856 & 9.00 & 295 & 1.93 & 0.00 & 0.00 & 4.66 & 55.00 & 5308 & 34.76 & 0.00 & 0.00 & 1.58 & 0.33 & 0.01 & 0.01 \\
  rno04660 & rno04660 & T cell receptor signaling pathway - Rattus norvegicus (rat) & 106 & 8856 & 11.00 & 295 & 3.53 & 0.00 & 0.01 & 3.12 & 79.00 & 5308 & 63.53 & 0.00 & 0.00 & 1.24 & 0.39 & 0.01 & 0.01 \\
  rno04657 & rno04657 & IL-17 signaling pathway - Rattus norvegicus (rat) &  95 & 8856 & 9.00 & 295 & 3.16 & 0.00 & 0.03 & 2.84 & 59.00 & 5308 & 56.94 & 0.37 & 0.46 & 1.04 & 0.36 & 0.01 & 0.01 \\
   \hline
\end{tabular}
\end{adjustbox}
\end{table}
